---
title: "OzoneXpect App"
author: "Akash Mer"
date: "`r Sys.Date()`"
output:
    slidy_presentation:
        keep_md: true
        theme: cerulean
        highlight: breezedark
knit: (function(input, ...){
    rmarkdown::render(input,
        output_dir = "C:/Users/akash/Documents/datasciencecoursera/OzoneXpect/docs",
        output_file = file.path("./index"))
    })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## **Introduction**

**[OzoneXpect](http://akashmer.shinyapps.io/OzoneXpect)** aims to **predict mean Ozone levels in parts per billion(ppb)** based on the measurement taken or the information known by the user.  
The user is allowed to input the following information :
  
1. **Measurement taken/information known to the user** - Represented as a predictor in the app, user can choose from the following,
    + **Temperature***(default)*
    + **Wind speed**
    + **Solar Radiation**
2. **Current Month** - Month the above measurement was taken in. Also gives the option to not specify the month. User can choose from **May to September** or **Any Month***(default)*
3. **Measurement Value** - User can then enter the measurement value to be used for the prediction.
  
**Data Used** - The data for the app comes from the `airquality` data set in R `datasets` package which is as follows,
```{r loadData, echo=TRUE, cache=TRUE, comment=""}
str(airquality)
```
**`r 100 - round(mean(complete.cases(airquality))*100, 2)`% of rows** contain missing values. `impute.knn()` function from the `impute` package was used to impute these mmissing values

## **How does it predict?**

**[OzoneXpect](http://akashmer.shinyapps.io/OzoneXpect)** predicts using 2 different models,
  
1. **Linear Model** - A linear model is built using the desired predictor as the outcome and using the data from the `airquality` data set. The data is subsetted in case a particular month was selected to ensure stratification and avoid any confounding due to the *month* variable. This model is built using the `lm()` function in R.
2. **Loess Model** - A non-linear model is built under the same circumstances as above. This model is built using the `loess()` function in R with a **span of 0.7**
  
Then, the mean ozone level is predicted using both models and returned with a 95% prediction interval for both models  

## **Outputs**
  
### **1. Relationship Plot**

Salient Features : 
  
* An interactive scatter plot between ozone and the desired predictor is displayed as soon as any **predictor and month is selected and changes based on the changes in those 2 inputs**.
* **Lines representing both the models** are also included with the option to click and select which one appears left to the user
* Once the measurement input is done, the plot is updated with the prediction points plotted as well

```{r examplePlot, cache=TRUE, warning=FALSE, message=FALSE}
if(system.file(package = "dplyr") == "") install.packages("dplyr")
if(system.file(package = "impute") == "") install.packages("impute")
if(system.file(package = "ggplot2") == "") install.packages("ggplot2")
if(system.file(package = "plotly") == "") install.packages("plotly")
library(dplyr)
library(impute)
library(ggplot2)
library(plotly)

# Loading the data set
# Imputing the missing values with rng seed 541258
# Converted to a matrix since that's the way impute.knn works
dat <- impute.knn(as.matrix(airquality), k = 10, rng.seed = 541258)$data

# Converting it back to a data frame
dat <- as.data.frame(dat) %>%
    # Rounding off the imputed values since ozone values should be integer
    mutate(Ozone = round(Ozone),
           # Same for Solar radiation
           Solar.R = round(Solar.R),
           # Converting month column to a factor
           Month = factor(Month, labels = c("May", "June", "July", "August",
                                                "September")))

# Changing the column names to lower case for easier typing and matching
names(dat) <- tolower(names(dat))

# Defining the Simple Linear Model for each predictor
modelLinear <- lm(ozone ~ temp, dat)
modelLoess <- loess(ozone ~ temp, dat, span = 0.7)

# Defining the plotting points for x axis
newDat <- data.frame(temp = seq(min(dat$temp), max(dat$temp), by = 0.05))

# Predicting on the linear model
linear <- predict(modelLinear, newdata = newDat)
            
# Predicting on the loess model
smooth <- predict(modelLoess, newdata = newDat)
    
# Prediction value for the linear model
valueDat <- data.frame(temp = 78)
predLinear <- predict(modelLinear, newdata = valueDat)
            
# Predicting value for the loess model
predLoess <- predict(modelLoess, newdata = valueDat)
            
# Adding to the data fram to display
valueDat <- cbind(valueDat, predLinear, predLoess)
            
# Creating the plot
plotModel <- ggplot() +
    # Plot the ozone against the predictor the user wanted
    geom_point(data = dat,
           aes(temp, ozone),
               size = 0.7) +
    # Plot the linear model line
    geom_line(data = newDat, aes(temp, linear,
                             color = "Linear Model"),
              linewidth = 1.5) +
    # Plot the loess model line
    geom_line(data = newDat, aes(temp, smooth,
                                 color = "Loess Model"),
              linewidth = 1.5) +
    # Adding the prediction points
    geom_point(data = valueDat, aes(temp, predLinear,
                             color = "Linear Model"), size = 3) +
    geom_point(data = valueDat, aes(temp, predLoess,
                             color = "Loess Model"), size = 3) +
    # Changing labels and adding title
    labs(title = "Example Plot", y = "Ozone (ppb)",
         x = "Temperature (°F)") +
    # Adding the legend for different models
    scale_color_manual(name = "Model",
                       values = c("#F8766D", "#00B0F6"),
                       breaks = c("Linear Model", "Loess Model")) +
    # Changing to black and white thee
    theme_bw()
            
# Returning a plotly object
ggplotly(plotModel)
```
  
### **2. Predictions**

Salient Features :
  
* A table is displayed to the user **only if the input values are within the corresponding bounds of minimum and maximum values** in the `airquality` data set to ensure higher accuracy.
* Any **negative values taken by the prediction or the 95% prediction interval bounds are highlighted with orange** to inform the user that these values are incorrect.
  
```{r examplePrediction, cache=TRUE, warning=FALSE, message=FALSE, dependson="examplePlot"}
if(system.file(package = "gt") == "") install.packages("gt")
library(gt)

# Predicting om the linear model
ozoneLinear <- predict(modelLinear, newdata = valueDat,
                                   interval = "prediction")

# Predicting on the loess model
predLoess <- predict(modelLoess, newdata = valueDat, se = TRUE)
ozoneLoess  <- data.frame(predLoess$fit,
                          predLoess$fit - predLoess$se.fit,
                          predLoess$fit + predLoess$se.fit)
            
# Combining the predictions into a single data framw
result <- data.frame(ozone=c(ozoneLinear[1,1], ozoneLoess[1,1]),
                     lower=c(ozoneLinear[1,2], ozoneLoess[1,2]),
                     upper=c(ozoneLinear[1,3], ozoneLoess[1,3]))
            
# Creating a gt table object for printing
tab <- gt(result) %>%
    tab_header(
        title = md("**Example Prediction**")) %>%
        tab_spanner(
        label = md("**95% Prediction Interval**"),
        columns = c("lower", "upper")
    ) %>%
    cols_label(
        ozone = md("**Mean Ozone Level(ppb)**"),
        lower = md("*lower limit*"),
        upper = md("*upper limit*")
    ) %>%
    tab_row_group(
        label = md("**Loess Model**"),
        rows = 2
    ) %>%
    tab_row_group(
        label = md("**Linear Model**"),
        rows = 1
    ) %>%
    tab_style(
        style = cell_fill(color = "orange", alpha = 0.5),
        locations = cells_body(columns = everything(),
                               rows = ozone < 0 |
                                   lower < 0 |
                                   upper < 0)
    ) %>%
    tab_footnote(
        footnote = md("*These negative values are not helpful*"),
        locations = cells_body(columns = everything(),
                               rows = ozone < 0 |
                                   lower < 0 |
                                   upper < 0)
    ) %>%
    tab_source_note(
    source_note = md("*Reference: airquality data set from R 
                     datasets package*")
    )

# Returning the table
tab
```

## **Strengths**
  
1. The user is **afforded a lot of input options** like deciding the predictor, informing the app about the current month, and thus is able to **get mean ozone levels depending on a variety of conditions**.
2. An **interactive** plot is displayed which **updates depending on the user's input**
3. **Both linear and non-linear predictions are returned**, thus allowing the user to choose whatever predictions they want
4. User is **informed about incorrect predictions as well**
5. Predictions are coupled with **95% prediction intervals** thus providing the probability statistic behind the uncertainty in the predictions
  
### *Limitations and Plans to tackle the limitations*
  
1. Small data set - In search for larger data sets representing similar variables
2. Only 6/12 months included - In search for data sets containing measurements for around the year
3. Data was not divided into train/test sets due to the already small sample size - Prediction intervals and the associated probability of error is provided though
  
### **[LINK TO THE APPLICATION](http://akashmer.shinyapps.io/OzoneXpect)**